---
title: "Getting Started with R: A Beginner's Guide"
author: "Rahmatullah khan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This tutorial will help you get started with R programming. We'll cover setting directories, creating simple plots, and basic data manipulation - perfect for your first steps in R!

# Setting Up Your Working Directory

## Understanding Working Directories

The working directory is the folder where R looks for files and saves outputs by default.

```{r check-directory}
# Check current working directory
getwd()

# List files in current directory
list.files()
```

## Setting Your Working Directory

```{r set-directory, eval=FALSE}
# Method 1: Using setwd() - adjust path to your folder
setwd("~/Documents/R_Tutorial/Getting_Started")  # Mac/Linux
# setwd("C:/Users/YourName/Documents/R_Tutorial")  # Windows
```

# Creating and Setting Output Directory

Let's create a folder for our outputs and set it as our working directory:

```{r create-output-dir}
# Create output directory if it doesn't exist
output_dir <- "outputs"
if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

# Check the directory was created
list.files()

# Set output directory (optional - you can save directly to it)
# setwd(output_dir)
```

# Simple Data Creation

Let's create some sample data to work with:

```{r create-data}
# Create simple data frame
student_data <- data.frame(
  student_id = 1:20,
  name = paste("Student", 1:20),
  math_score = c(85, 92, 78, 88, 95, 76, 82, 90, 87, 79,
                 93, 81, 89, 77, 94, 86, 83, 91, 75, 84),
  science_score = c(88, 85, 82, 90, 92, 79, 85, 88, 91, 83,
                    89, 80, 87, 78, 93, 84, 86, 89, 77, 85),
  grade = sample(c("A", "B", "C"), 20, replace = TRUE)
)

# View the data
head(student_data)
```

# Creating Simple Plots

## Basic Scatter Plot

```{r scatter-plot}
# Simple scatter plot
plot(student_data$math_score, student_data$science_score,
     main = "Math vs Science Scores",
     xlab = "Math Score",
     ylab = "Science Score",
     pch = 16,
     col = "blue")

# Add a trend line
abline(lm(science_score ~ math_score, data = student_data), 
       col = "red", lwd = 2)
```

## Bar Plot

```{r bar-plot}
# Count of students by grade
grade_counts <- table(student_data$grade)

# Create bar plot
barplot(grade_counts,
        main = "Student Distribution by Grade",
        xlab = "Grade",
        ylab = "Number of Students",
        col = c("lightblue", "lightgreen", "lightcoral"))
```

## Histogram

```{r histogram}
# Histogram of math scores
hist(student_data$math_score,
     main = "Distribution of Math Scores",
     xlab = "Math Score",
     ylab = "Frequency",
     col = "lightblue",
     border = "black",
     breaks = 8)
```

## Box Plot

```{r box-plot}
# Box plot comparing scores by grade
boxplot(math_score ~ grade, data = student_data,
        main = "Math Scores by Grade",
        xlab = "Grade",
        ylab = "Math Score",
        col = c("lightblue", "lightgreen", "lightcoral"))
```

# Saving Plots to Output Directory

```{r save-plots}
# Set up output directory for saving
output_path <- "outputs"

# Save scatter plot
png(file.path(output_path, "scatter_plot.png"), width = 600, height = 400)
plot(student_data$math_score, student_data$science_score,
     main = "Math vs Science Scores",
     xlab = "Math Score",
     ylab = "Science Score",
     pch = 16,
     col = "blue")
abline(lm(science_score ~ math_score, data = student_data), 
       col = "red", lwd = 2)
dev.off()

# Save bar plot
png(file.path(output_path, "grade_distribution.png"), width = 600, height = 400)
grade_counts <- table(student_data$grade)
barplot(grade_counts,
        main = "Student Distribution by Grade",
        xlab = "Grade",
        ylab = "Number of Students",
        col = c("lightblue", "lightgreen", "lightcoral"))
dev.off()

# Check what was saved
list.files(output_path)
```

# Basic Data Analysis

```{r basic-analysis}
# Summary statistics
summary(student_data[, c("math_score", "science_score")])

# Calculate correlations
cor(student_data$math_score, student_data$science_score)

# Average scores by grade
aggregate(cbind(math_score, science_score) ~ grade, 
          data = student_data, mean)
```

# Tips for First-Time R Users

## Essential R Commands

```{r essential-commands}
# View data
head(student_data)      # First 6 rows
tail(student_data)      # Last 6 rows
str(student_data)       # Structure of data
names(student_data)     # Column names

# Install and load packages
# install.packages("ggplot2")  # Install package
# library(ggplot2)             # Load package
```

## File Operations

```{r file-operations}
# Save data
write.csv(student_data, file.path(output_path, "student_data.csv"), row.names = FALSE)

# Read data (example)
# my_data <- read.csv("data/my_file.csv")

# List files
list.files()
list.files(output_path)
```

# Conclusion

Congratulations! You've learned the basics of:

- Setting and managing working directories
- Creating and manipulating simple data
- Making various types of plots
- Saving outputs to specific directories
- Basic data analysis

## Next Steps

1. Try modifying the plots with different colors and labels
2. Create your own data sets
3. Explore more plot types with `ggplot2`
4. Learn about data manipulation with `dplyr`

## Directory Structure After Running This Tutorial

After running this code, your project folder should look like this:

```
your_project/
├── your_r_markdown_file.Rmd
├── outputs/
│   ├── scatter_plot.png
│   ├── grade_distribution.png
│   └── student_data.csv
└── README.md (if you have one)
```

Happy coding with R! 🚀